\input{include.tex}

\markright{Piotr Suwara\hfill Rachunek prawdopodobieństwa II: 27 listopada 2013 \hfill}

\begin{document}
	\begin{definition}[filtration]
		$(\F_t)_T$, for $T$ a~segment in $\mb{Z}$, is a~\emph{filtration},
		if $\F_t \subset \F$ is a~$\sigma$-body and $\forall_{t \leq s} \F_t \subset \F_s$.
	\end{definition}
	
	\begin{definition}[stopping time]
		$\tau: \Omega \to T \cup \{\infty\}$ is a~\emph{stopping time}, if
		$\forall_{t \in T} \{\tau \leq t\} \in \F_t$
		($\iff \forall_{t \in T} \{ \tau = t\} \in \F_t\}$).
	\end{definition}
	
	\begin{definition}
		Let $(\F_t)$ be a filtration, $\tau$ be a~stopping time, then define
		$$\F_\tau = \{ A \in \F: A \cap \{ \tau \leq t\} \in \F_t\}.$$
	\end{definition}
	
	\begin{proposition}
		$\F_\tau = \{ A \in \F: A \cap \{ \tau = t \} \in \F_t\}$.
	\end{proposition}
	
	\begin{proposition}
		$\tau_1, \tau_2$ stopping times, then $\tau_1 \wedge \tau_2 = \min(\tau_1, \tau_2)$
		and $\tau_1 \vee \tau_2 = \max(\tau_1, \tau_2)$ are too.
		
		$\tau = t$ is a~stopping time.
		
		$\tau_1 \leq \tau_2$ stopping times $\implies \F_{\tau_1} \subset \F_{\tau_2}$
		
		$\tau$ jest $\F_\tau$-mierzalne.
	\end{proposition}
	
	\begin{definition}[adapted process]
		$(X_t)_{t \in T}$ is \emph{adapted to the filtration} $(\F_t)_{t \in T}$ or just \mbox{\emph{$(\F_t)$-adapted}}, if $\forall_t$ $X_t$ is $\F_t$-measurable.
	\end{definition}
	
	\begin{proposition}
		$(\F_t)$ filtration, $(X_t)$ is $(\F_t)$-adapted, $\tau$ a~stopping time,
		then $\tau < \infty \implies$ \mbox{$X_\tau$ is $\F_\tau$-measurable}.
		
		More generally, $X_\tau$ is $\F_\tau$-measurable on the set $\{\tau < \infty\}$, \\ \mbox{
		i.e. $\forall_{B \in \B(\R)} \{X_\tau \in B\} \cap \{t < \infty\} \in \F_\tau$.}
	\end{proposition}
	
	\pagebreak
	
	\begin{definition}[martingale]
		$(X_t)$ is a~\emph{martingale} (resp. submartingale, supermartingale)
		with respect to a~foltration $(\F_t)$, if 
		\begin{itemize}
			\item $\forall_{t \in T}$ $X_t$ is $\F_t$-measurable,
			\item $\forall_{t \in T} \E |X_t| < \infty$,
			\item $\forall_{s \leq t, s, t \in T} \E(X_t | \F_s) = X_s$ a.s. (resp. $\geq$, $\leq$).
		\end{itemize}
	\end{definition}
	
	\begin{remark}
		$X_t$ is a~martingale iff it is both a~submartingale and a~supermartingale.
	\end{remark}
	
	\begin{remark}
		$(X_t)$ is a~$(\F_t)$-martingale if $X_t$ is $\F_t$-measurable, integrable and \\ \mbox{$\forall_{s < t, A \in \F_s} \E(X_s \1_A) = \E(X_t \1_A)$ (resp. $\leq$ for submartingale, $\geq$ for supermartingale).}
	\end{remark}
	
	\begin{remark}
		For $T$ a~segment in $\mb{Z}$, $(X_t)$ is $(\F_t)$-martingale iff
		$X_t$ is $\F_t$-measurable, $\E|X_t|<\infty$, $\E(X_{s+1} | \F_s) = X_s$ a.s. (resp. $\geq$ for submartingale, $\leq$ for supermartingale).
	\end{remark}
	
	\begin{remark}
		$X_t$ submartingale iff $-X_t$ supermartingale.
	\end{remark}
	
	\begin{remark}
		$X_t, Y_t$ are $\F_t$-martingales, then $a X_t + b Y_t$ also (for submartingale take $a, b \geq 0$).
	\end{remark}
	
	\begin{definition}
		$(\F_n)_{n \geq 0}$ filtration generated by $(X_1, X_2, \ldots)$, $\F_0 = \{ \varnothing, \Omega\}, \F_n = \sigma(X_1, \ldots, X_n)$.
	\end{definition}
	
	\begin{fact}
		$X_1, X_2, \ldots$ independent random variables, $S_0 = 0$, $S_n = X_1 + \ldots + X_n$, \mbox{$(\F_n)$ filtration} generated by $(X_n)$.
		
		Then $S_n$ is a~martingale iff $X_n$ are integrable and $\E X_n = 0$ ($\geq$ for submartingale, \mbox{$\leq$ for supermartingale}).
	\end{fact}
	
	\begin{fact}
		$X$ integrable random variable, $(\F_t)$ filtration, $X_t = \E(X | \F_t)$ is a~$(\F_t)$-martingale.
	\end{fact}
	
	\begin{fact}
		$(X_t)$ is a~$(\F_t)$-martingale, $\varphi:\R \to \R$ convex, $\E |\varphi(X_t)| < \infty$, then $(\varphi(X_t), \F_t)$ is a~submartingale.
	\end{fact}
	
	\begin{corollary}
		$(X_t, \F_t)$ a~martingale, $p \geq 1$, $\E |X_t|^p < \infty$, then $(|X_t|^p, \F_t)$ is a~submartingale.
	\end{corollary}
	
	\begin{corollary}
		$(X_t, \F_t)$ submartingale, then $(X_t \vee a, \F_t)$ submartingale.
	\end{corollary}
	
	\begin{corollary}
		$(X_t, \F_t)$ martingale, then $(X_t^+, \F_t)$ and $(X_t^-, \F_t)$ submartingales (where $Y^+ = Y \wedge 0$, $Y^- = (-Y) \wedge 0$).
	\end{corollary}
	
	\begin{fact}[martingale transformation]
		$(X_n, \F_n)$ martingale, let $Y_n = X_0 + V_1(X_1 - X_0) + V_2 (X_2 - X_1) + \ldots + V_n (X_n - X_{n-1})$ for $V_n$ being $(\F_{n-1})$-measurable and bounded, then $(Y_n, \F_n)$ is a~martingale.
	\end{fact}








\end{document}
 
 
